\documentclass[11pt,a4paper]{article}

\input{../shared/macros}

\hypersetup{
  colorlinks = true,
  linkcolor  = blue!60!black,
  citecolor  = green!50!black,
  urlcolor   = blue!60!black,
  pdftitle   = {Phase 2 Timing Profile},
  pdfauthor  = {surface-partition project}
}

\title{%
  \textbf{Phase~2 IPOPT Callback Timing Profile}\\[0.5em]
  \large Empirical Cost Breakdown of the Perimeter-Refinement Solver
}
\author{}
\date{May 2026}

\begin{document}
\maketitle

\begin{abstract}
This document records empirical timing measurements collected by the
\code{ProfilingState} instrumentation in \codefile{src/profiling.py}.
Each row of Table~\ref{tab:reference-breakdown} reports the wall-clock
cost and the Steiner-recomputation count of one IPOPT callback
(objective, gradient, constraints, Jacobian, Hessian) for a single
Phase~2 run on a 10-cell torus.  The data identifies the Hessian
callback --- specifically the finite-difference Steiner contribution
documented in \S{}6 of \codefile{01-phase2-derivatives} --- as the sole
performance bottleneck, consuming $98.41\%$ of total wall time.  This
document is a \emph{living artefact}: Section~\ref{sec:scaling} will be
extended as profiles for other mesh resolutions and cell counts are
collected.
\end{abstract}

\tableofcontents
\newpage

% =========================================================================
\section{Overview}
% =========================================================================

Phase~2 of the \codefile{surface-partition} pipeline solves the
constrained nonlinear program~\eqref{eq:nlp} using IPOPT
\cite{wachter2006ipopt}.  At each interior-point iteration IPOPT
invokes five callbacks:

\begin{center}
\renewcommand{\arraystretch}{1.2}
\begin{tabular}{ll}
\toprule
\textbf{Callback} & \textbf{Quantity returned} \\
\midrule
\texttt{objective}   & perimeter value $P(\Lam)$ \\
\texttt{gradient}    & perimeter gradient $\grad P(\Lam)$ \\
\texttt{constraints} & area vector $(A_0,\ldots,A_{N-2})(\Lam)$ \\
\texttt{jacobian}    & area Jacobian $J(\Lam)$ \\
\texttt{hessian}     & Lagrangian Hessian $H = \objfac\,\Hess{P} + \sum_k \mu_k\,\Hess{A_k}$ \\
\bottomrule
\end{tabular}
\end{center}

\noindent
The derivations of all five quantities are given in document
\codefile{01-phase2-derivatives}; the present document reports only
their \emph{measured cost}.  File I/O (HDF5 checkpoint reads and writes), topology migrations, and
HDF5 checkpointing are excluded from all figures reported here.

\begin{equation}
  \min_{\Lam \in \mathbb{R}^n}\; P(\Lam)
  \quad\text{subject to}\quad
  A_k(\Lam) = \Atgt,\quad k = 0,\ldots,N-2
  \label{eq:nlp}
\end{equation}

% =========================================================================
\section{Profiling Methodology}
\label{sec:method}
% =========================================================================

\subsection{What is recorded}

The class \code{ProfilingState} (\codefile{src/profiling.py}) accumulates
two quantities per callback type across all IPOPT iterations and all
topology cycles within a single run:

\begin{itemize}[noitemsep]
  \item \textbf{Wall-clock time} --- total and per-call mean for each
        callback.
  \item \textbf{Steiner recomputation count} --- a running integer
        incremented each time \code{record\_steiner()} is called inside a
        callback.  One increment corresponds to one full evaluation of the
        Steiner geometry (Fermat-Torricelli point and associated perimeter
        or area) for all active triple points.  This counter directly
        measures the number of finite-difference perturbations executed
        within a callback, making it the primary diagnostic for the FD
        cost.
\end{itemize}

Results are written to \code{timing\_profile.yaml} in the campaign
directory when \texttt{-{}-profile} is passed to
\codefile{scripts/refine\_perimeter.py}.

\begin{tcolorbox}[colback=blue!4!white, colframe=blue!40!black,
                  title={Code: profiling accumulator}, fontupper=\small]
\codefile{src/profiling.py}: \code{ProfilingState}\\[2pt]
\code{record(key, dt)} --- accumulate wall time.\\
\code{record\_steiner(key, n)} --- increment the Steiner recomputation count.\\
\code{finalize()} --- compute means and percentage breakdown.\\
\code{to\_yaml\_dict()} --- serialize to \code{timing\_profile.yaml}.
\end{tcolorbox}

\subsection{Why only Jacobian and Hessian report recomp counts}

All Steiner derivative quantities --- the perimeter gradient, the area
Jacobian, and both Hessians --- are computed by finite differences (see
document \codefile{01-phase2-derivatives}, \S{}6).  None of these
quantities has an analytical derivative implementation.  However, the
Steiner recomputation count is only recorded for the Jacobian and
Hessian callbacks.  This is an instrumentation choice: the profiling
handle \code{\_prof} is threaded into the vectorized Steiner functions in
\codefile{src/partition/vectorized\_steiner.py} only when they are called
from those two paths.  In the gradient callback, the same finite-difference
loop executes and the Steiner geometry is re-evaluated for every
Steiner-affected VP, but \code{\_prof} is not passed in, so those
evaluations are not counted.  The gradient's Steiner FD cost is real but
small (absorbed into the $0.54\%$ wall time in
Table~\ref{tab:reference-breakdown}); the counter was wired primarily
where the cost was expected to be large.

\begin{remark}[FD structure and recomp counts]
A single gradient call perturbs each of the $\sim\!180$ Steiner-affected
VPs once (forward FD, $\varepsilon = 10^{-6}$) and evaluates the Steiner
geometry at each perturbation --- but these are not counted.  A single
Jacobian call does the same for the area constraints, and \emph{is}
counted: $\sim\!61$ evaluations for a run with 20 triple points.  A
single Hessian call then re-runs the full FD gradient and Jacobian once
per Steiner-affected VP (the outer FD layer), giving
$180 \times 61 \approx 11\,000$ counted evaluations per Hessian call ---
in agreement with the measured 11\,041.
\end{remark}

% =========================================================================
\section{Reference Run}
\label{sec:ref-run}
% =========================================================================

All numbers in Section~\ref{sec:results} come from a single baseline
run whose parameters are given in Table~\ref{tab:reference-run}.  This
run used the exact Hessian option (\texttt{exact\_hessian: true}) and
\texttt{allow\_partial\_convergence: true}, which allows the pipeline to
proceed to topology migration even when IPOPT exits with a non-fatal
status code.

\begin{table}[H]
\centering
\caption{Reference run parameters.}
\label{tab:reference-run}
\renewcommand{\arraystretch}{1.2}
\begin{tabular}{ll}
\toprule
\textbf{Parameter} & \textbf{Value} \\
\midrule
Surface             & torus ($R=1$, $r=0.4$) \\
Number of cells $N$ & 10 \\
Mesh vertices       & 242\,608 \\
Active VPs          & 7\,110 \\
Triple points       & 20 \\
Solver              & IPOPT, exact Hessian \\
IPOPT iterations    & 369 (across 2 topology cycles) \\
Wall-clock total    & 1\,284.32\,s \\
Campaign directory  & \small\texttt{ipopt\_btol0.001\_lbfgs30\_hess\_bestiter\_partial} \\
\bottomrule
\end{tabular}
\end{table}

% =========================================================================
\section{Results}
\label{sec:results}
% =========================================================================

Table~\ref{tab:reference-breakdown} gives the full per-callback
breakdown extracted from \code{timing\_profile.yaml} for the reference
run of Section~\ref{sec:ref-run}.

\begin{table}[H]
\centering
\caption{Per-callback timing breakdown, reference run ($N=10$, 242\,608 vertices).}
\label{tab:reference-breakdown}
\renewcommand{\arraystretch}{1.25}
\begin{tabular}{lrrrrr}
\toprule
\textbf{Callback} & \textbf{Calls} & \textbf{Total (s)} &
  \textbf{\% wall} & \textbf{Mean (s/call)} & \textbf{Steiner recomps/call} \\
\midrule
Objective    & 1\,724 &     1.00  &  0.08          & 0.0006 & --- \\
Gradient     &   372  &     6.95  &  0.54          & 0.019  & --- \\
Constraints  & 1\,724 &     2.46  &  0.19          & 0.0014 & --- \\
Jacobian     &   372  &     8.26  &  0.64          & 0.022  & 61 \\
Hessian      &   368  & 1\,263.95 & \textbf{98.41} & 3.43   & 11\,041 \\
\midrule
\textit{Total} & --- & 1\,284.32 & 100.00 & --- & 4\,085\,780 \\
\bottomrule
\end{tabular}
\end{table}

% =========================================================================
\section{Analysis: the Hessian Bottleneck}
\label{sec:hessian-bottleneck}
% =========================================================================

The Hessian callback dominates runtime by a factor of $\sim\!150$ over
the next most expensive callback (Jacobian at $0.64\%$).  The cause is
the nested finite-difference scheme used for the Steiner contributions
$\Hess{\Pst}$ and $\Hess{A^{\mathrm{st}}_k}$, described in \S{}6.4
and \S{}6.5 of document \codefile{01-phase2-derivatives}.

\paragraph{Cost decomposition.}
Let $T$ denote the number of active triple points and $n = |S|$ the
total number of Steiner-affected VPs (the set \code{tp\_affected\_vps}
in \codefile{src/partition/partition\_arrays.py}).  Each triple-point
triangle has exactly 3 boundary edges, each carrying one VP, so
$n = 3T$.  In the reference run $T = 20$, giving $n = 60$.

A single Jacobian call evaluates the Steiner geometry once as a base,
then once per VP in $S$, for a total of $n + 1$ recomputations --- in
agreement with the measured 61.

The Hessian executes two nested FD loops.  The perimeter Hessian
(\S{}6.4 of \codefile{01-phase2-derivatives}) uses central differences:
for each of the $n$ VPs in $S$ it calls the gradient FD twice
($+\varepsilon$ and $-\varepsilon$), and each gradient call costs $n+1$
recomputations.  The area Hessian (\S{}6.5) uses forward differences:
one base Jacobian call plus $n$ perturbed Jacobian calls, each costing
$n+1$ recomputations.  The total is

\begin{equation}
  \text{recomps per Hessian call}
  \;=\;
  \underbrace{2n(n+1)}_{\Hess{\Pst}}
  \;+\;
  \underbrace{(n+1)^2}_{\Hess{A_k^{\mathrm{st}}}}
  \;=\;
  (n+1)(3n+1)
  \label{eq:hess-cost}
\end{equation}

\noindent
With $n = 60$: $(61)(181) = 11\,041$ --- in exact agreement with the
measured value.

\begin{remark}[Quadratic scaling with triple-point count]
Since $n = 3T$, Equation~\eqref{eq:hess-cost} gives
$(n+1)(3n+1) \approx 3n^2 = 27T^2$ for large $T$.  The Hessian cost
therefore scales \emph{quadratically} with the number of triple points.
Increasing the number of cells $N$ increases $T$ and therefore the cost
grows as $\mathcal{O}(T^2)$.  This is the key motivation for implementing
an analytical Steiner Hessian.
\end{remark}

\paragraph{Why the exact Hessian is required.}
An L-BFGS quasi-Newton approximation could avoid the FD Hessian cost
entirely, but in practice the L-BFGS approximation does not converge
tightly enough to push VPs to the mesh boundary, so topology migrations
are never triggered.  The exact Hessian is therefore required for
production runs that need to discover the correct partition topology.

% =========================================================================
\section{Scaling Outlook}
\label{sec:scaling}
% =========================================================================

This section will be extended as additional runs are profiled.  Two
scaling axes are planned:

\begin{itemize}[noitemsep]
  \item \textbf{Mesh refinement.}  Fix $N=10$, vary the torus mesh
        resolution $(n_\theta, n_\phi)$, and record
        Table~\ref{tab:reference-breakdown}-style breakdowns.
  \item \textbf{Cell-count scaling.}  Fix a moderate mesh resolution,
        vary $N$, and measure how the number of triple points and the
        Hessian recomp count scale with $N$.
\end{itemize}

% Tables and figures will be inserted here as profile data becomes available.

% =========================================================================
\section*{Summary table}
% =========================================================================

\begin{center}
\renewcommand{\arraystretch}{1.35}
\begin{tabular}{p{3.8cm}p{4.5cm}p{4.5cm}}
\toprule
\textbf{Callback} & \textbf{Current cost (reference run)} &
  \textbf{Planned optimization} \\
\midrule
Objective, constraints
  & $<0.3\%$ of wall; negligible
  & none \\
Gradient, Jacobian
  & $\sim\!1\%$ of wall; fast
  & none \\
Hessian
  & $98.41\%$ of wall; FD Steiner dominates
  & Analytical Steiner Hessian (see
    \codefile{01-phase2-derivatives} \S{}6 for the FD baseline) \\
\bottomrule
\end{tabular}
\end{center}

\bibliographystyle{plain}
\bibliography{../shared/references}

\end{document}
